Increasing User Engagement in a Volunteer-Based Ephemeral Evolutionary Computation System

### 2017-11-08 2017-EUROPAR/Gamification/paper.tex
Some edits

Distributed computing systems can be created using volunteers who decide to
provide their own  resources or storage to contribute to a common effort. They can, for instance, run a script embedded in a web page, collaboration is straightforward, but also ephemeral, with resources depending on the amount of time a user lends; this  implies that 
the user has to be kept engaged so as to obtain as many computing cycles as
possible. In this paper, we analyze a volunteer-based evolutionary computing system called
% NodIO
Anonymous with the objective of discovering rules that encourage volunteer
participation thus increasing the overall computing power. We present the results of
an experiment where a gamification technique is applied by adding a leader-board
showing the top scores achieved by registered contributors. In the
% NodIO
Anonymous system volunteers can
participate without the need to create an account, so the question was
if the need to register would have a negative impact on user participation.
The experiment results show that even if only a small percentage of users created an account,
those participating in the competition provided around 90\% of the work.

@@

There are several open issues in the research regarding the application of volunteer computing techniques to the development of distributed evolutionary algorithms of which the most important is approaching the problem as a socio-technical system \cite{vespignani2009predicting}. which integrates user decisions and behavioral patterns in the model; this
includes trying to optimize the number of users in a particular
experiment. The challenge is to design a system that, whatever the
number of users available to perform the experiment, is able to
maximize their contribution to the evolutionary algorithm. One obvious
way of achieving that is trying to find as many users as possible. But
another approach, and the one that we will be following in this paper, is
to delve into the socio-technical nature of the system, using
gamification techniques to improve the amount of time every user will lend to the system. Gamification has to include some way of user authentification,
which is a hurdle that might take users away; however, the open design
of the system we use means that we can actually mix anonymous-non
gamified and authentified-gamified users, with only the small overhead
of adding the gamification mechanism.

There is a trade-off between ease of use, where users just need to visit a web page, versus personlaization, where users need to register and subsecuently authenticate in order to participate. Registration could be seen as a barrier to volunteers trying to participate in an ephemeral experiment, so the question is what is the impact on user participation.

The rest of the paper is organized as follows: Next we will briefly
present the state of the art in opportunistic distributed evolutionary
computation (EC). Section \ref{sec:gamification} will describe the
framework and problem used in the experiments, which are publicly
available under a free license. We will present the results of the
different steps in the incremental design in Section
\ref{sec:experiments}, to finally wrap up with the conclusions.

@@

Volunteer computing involves users deciding to run a program that acts
as a client or as a peer in a distributed computing experiment
and, as such, has been deployed in many different ways from the
beginning of the Internet, starting with the SETI@home framework for
processing extraterrestrial signals \cite{david-seti:home}, or a
high-throughput queuing system such as HTCondor \cite{HTCondor}.
However,
the dual facts of the introduction of JavaScript as a universal language for the
browser and the browser itself as both an ubiquitous web and Internet client has
made this combination the most popular for volunteer computing
frameworks such as the one we are using here, and whose first version
was described in \cite{DBLP:conf/gecco/GuervosG15}. Systems based on the
JavaScript/browser combination
emphasize the ephemerality, ease of use, and universality, while
systems such as HTCondor or BOINC might be more adequate for work that % citation for BOINC (?)
require higher availability of volunteer resources, achieving this
with downloadable clients which in some cases, like HTCondor, need
administration access to the resources.

Several authors have already described systems using JavaScript either
for unwitting
\cite{unwitting-ec,boldrin2007distributed,apolonia2012enhancing} or volunteer
\cite{langdon:2005:metas,gecco07:workshop:dcor} distributed
evolutionary computation and it has been used ever since by several
authors, including more recent efforts
\cite{duda2013distributed,DBLP:journals/corr/abs-0801-1210,EvoStar2014:jsEO,martinez2015capataz,pan2015gray}. In fact, this last paper \cite{pan2015gray} performs an analysis of what is called {\em Gray computing} covering aspects of
feasibility, cost-effectiveness, change in the user experience and the
architectural optimization needed, concluding that the computing power
available is vast and it can be cost-effective to use it.

Using a peer to peer approach, the SPACE framework \cite{leclerc2016seamless} distributes fitness evaluations across a heterogeneous pool of cloud compute nodes and volunteer peer computers running a browser. In SPACE, peers establish a bi-directional communication with the server using the \texttt{Socket.io} JavaScript library.
As an experiment they provide an \texttt{ASM.js} compiled version of the RoboGen software platform, showing that JavaScript can be used in a broad
spectrum of applications.

Recent works have been using crowdsourcing in order to train robots
for human interaction, a representative work of this trend is that of
Anetsberger \& Bongard \cite{anetsbergerrobots} in which they propose
training simulated robots for the grounding of language symbols. They do this by using human observers that issue arbitrary commands to these robots via the web, providing positive or negative reinforcement in response to the
robot resulting actions. For interaction, they use the Twitch streaming platform where users can issue commands through the
platform's integrated chat service. Subjects were incentivized to
interact with the system by GUI features that provided participants
with a sense of involvement with the simulation. This kind of system
could add additional dynamics, because users can collaborate or
even compete with each other when issuing their commands. This approach has been called {\em social cloud} \cite{6404452}, emphasizing the fact that user participation is gamified in order to obtain the most from it \cite{7027564}.

The proof-of-concept efforts described above, do not go any further than  trying to find out how many users join the effort, and how many of them the system is able to support. In fact, systems such as the one described in
\cite{gecco07:workshop:dcor} had serious scaling issues; some of them
also tried to find out how much time was needed to find the solution
or, alternatively, how many users would be needed to be competitive
against single-user single-computer implementations of the same
algorithm. Lately, researchers have tried to integrate volunteer computing techinques as a part of a larger distributed computing effort \cite{leclerc2016seamless}. In fact, exclusive volunteer-based systems exhibit a certain amount of unpredictability and might be best used in combination with ready-available computing power.

As indicated in the introduction, the human is an integral part of the system, which can be considered as {\em human computation} \cite{quinn2011human}. Giving the user not only more control, by including registration, but also a measure to compare with other users might have positive effects. In fact, the BOINC projects applies a gamification technique
by having a web page to present the ``Top 100 multi-project BOINC participants'' where
the name of the volunteer, the number of projects, GFLOPS, country and team are presented
(\url{https://boinc.berkeley.edu/chart_list.php}).

@@

\section{Gamification}
\label{sec:gamification}
A definition given by Huotari  \cite{huotari2012defining} is ``Gamification is
the process by which gaming concepts are brought to the real world tasks associated with
real people''. Gamification uses game design elements out of the domain of games
with the objective of enhancing the user's experience, engagement, productivity,
learning, among others. Deterding et al. proposes the following definition:
 ``Gamification” is the use of game design elements in non-game contexts''
 \cite{deterding2011gamification}.

Gamification techniques in a volunteer context seeks to persuade
users to use their natural desire to compete, learn and socialize in
given non-game context application \cite{deterding2011game,hamari2014does}.
Some works give a form of reward to users, these include
points \cite{sutter2010browse}, achievement badges or levels \cite{hamari2011framework},
the filling of a progress bar \cite{o2010get}, or providing the user with virtual currency.
By Making the rewards for  tasks achievements visible to other players or
providing leader boards are ways of encouraging players to compete \cite{hickman2010total}. In this work, a leader board was implemented in order to promote competition, presenting only the all-time top five participating users. Competition can also have problematic consequences, which can result in
negative conduct, low cooperation and collaboration, or disadvantaging certain player demographics
such as women \cite{kumar2013gamification}. Another techniques to gamification
is to make existing tasks feel more like games \cite{deterding2010just}.
Some techniques used in this approach include adding meaningful choice,
on-boarding with a tutorial, increasing challenge, and adding narrative \cite{mcgonigal2011reality}.

@@

In this experiment a gamification technique is applied to
the Anonymous
% {\sf NodIO}
volunteer computing framework, using the
particular version described in \cite{DBLP:conf/gecco/MereloCGCRV16,2016arXiv160101607M}.
The objective of the experiment is to test the kind of impact
applying a gamification technique has on user engagement in this
particular application. In order to apply a rewarding system
user authentication had to be developed first. In earlier designs
this functionality was not desired because it was seen as a barrier
for participation. After all, the advantage of a browser based volunteer system
is precisely the minimum amount of user intervention needed to start.
The last thing a user wants to see is yet another registration form.
Then, the first design decision for this version is that registration is optional and simple as shown in Figure \ref{fig:login}.

@@

After login, registered users are welcomed by their name, and they can see
their score and level. All users can see in a modal dialog the leader board.
While the modal is opened, the scores are refreshed every second. All other
functionality is available to all users and is the same as the previous
%{\sf NodIO}
Anonymous version, showing the state of the current algorithm.
In the front page a link to the open source code of the experiment was also
available \url{https://undisclos.ed}
% \url{https://github.com/lucero21/login-master}

\subsection{Problem}
In the browser, each page visiting the experiment loads an HTTP Web Worker
that runs a local island of an evolutionary algorithm to solve a
multi-modal problem called {\em l-trap}, represented in figure \ref{fig:trap}, which has been used extensively
as a benchmark for evolutionary algorithms \cite{fernandes2009using,nijssen2003analysis}.
This function counts the number of bits in a sequence $l$ and assigns
the local maximum $a$ if it has got 0 bits and the global maximum $b$ if it has $l$
bits, this makes the  the fitness fall into a {\em trap}
as the number of bits is increased, decreasing linearly until a change in slope
is reached at point $z$, adding deceptive component for evolutionary algorithms.
To increase the difficulty trap functions can be concatenated.
In our case we have used $40$ concatenated traps. The trap function  defined as:
\[ f(u)=
    \begin{cases}
      \frac{a}{z}(z-u) & \text{if } u\leq z\\
      \frac{b}{l-z} (u-z)& \text{otherwise}
   \end{cases}
 \]
 %
\begin{figure*}[htb]
    \centering
        \includegraphics[width=2.5in]{img/trap.png}
    \caption{Trap function.
    }
    \label{fig:trap}
\end{figure*}

@@

@@

The objective of this work was to test the kind of impact
applying a gamification technique has on user engagement in
the Anonymous
% {\sf NodIO}
volunteer computing framework; In concordance with
the results obtained in other
browser-based volunteer systems, after
applying the gamification techniques, user participation was increased.

>>>>

Distributed computing systems can be created using volunteers who
provide their own computing resources or storage to contribute to a common effort. They can, for instance,  run a script embedded in a web page, collaboration is straightforward, but also ephemeral, with resources depending on the amount of time a user lends; this  implies that
the user has to be kept engaged so as to obtain as many computing cycles as
possible. In this paper, we analyze a volunteer-based evolutionary computing system called
NodIO with the objective of discovering rules that encourage volunteer
participation thus increasing the overall computing power. We present the results of
an experiment where a gamification technique is applied by adding a leader-board
showing the top scores achieved by registered contributors. In the NodIO system volunteers can
participate without the need to create an account, so the question was
if the need to register would have a negative impact on user participation.
The experiment results show that even if only a small percentage of users created an account,
those participating in the competition provided around 90\% of the work.

@@

There are several open issues in the research regarding the application of volunteer computing techniques to the development of distributed evolutionary algorithms of which the most important is approaching the problem as a socio-technical system \cite{vespignani2009predicting} which integrates user decisions and behavioral patterns in the model; this
includes trying to optimize the number of users in a particular
experiment. The challenge is to design a system that, whatever the
number of available users to perform the experiment, it is able to
maximize their contribution to the evolutionary algorithm. One obvious
way of achieving that is trying to find as many users as possible. But
another approach, and the one that we will be following in this paper, is
to delve into the socio-technical nature of the system, using
gamification techniques to improve the amount of time every user will lend to the system. Gamification has to include some way of user authentification,
which is a hurdle that might take users away; however, the open design
of the system we use means that we can actually mix anonymous-non
gamified and authentified-gamified users, with only the small overhead
of adding the gamification mechanism.

There is a trade-off between ease of use, where users just need to visit a web page, versus personalization, where users need to register and subsequently authenticate to participate. Registration can be seen as a barrier to volunteers trying to join an ephemeral experiment, so the question is: what is the impact on user participation.

The rest of the paper is organized as follows: Next we will briefly
present the state of the art in opportunistic distributed evolutionary
computation (EC). Section \ref{sec:gamification} will describe the
framework and the problem used in the experiments, which are publicly
available under a free license. The results of each of the steps in the incremental design are presented in Section
\ref{sec:experiments}, to finally wrap up with the conclusions.

@@

Volunteer computing involves users who decide to run a program that acts
as a client or as a peer in a distributed computing experiment
and, as such, has been deployed in many different ways from the
beginning of the Internet, starting with the SETI@home framework for
processing extraterrestrial signals \cite{david-seti:home}, or a
high-throughput queuing system such as HTCondor \cite{HTCondor}.
However,
the dual facts of the introduction of JavaScript as a universal language for the
browser and the browser itself as both an ubiquitous web and Internet client has
made this combination the most popular for volunteer computing
frameworks such as the one we are using here, and whose first version
was described in \cite{DBLP:conf/gecco/GuervosG15}. Systems based on the
JavaScript/browser combination
emphasize the ephemerality, ease of use, and universality, while
systems such as HTCondor or BOINC might be more adequate for work that % citation for BOINC (?)
requires higher availability of volunteer resources. In order to increase availability, systems rely on downloadable clients, and in some cases like HTCondor, they also require administrative access to certain resources.

Several authors have already described systems using JavaScript either
for unwitting
\cite{unwitting-ec,boldrin2007distributed,apolonia2012enhancing} or volunteer
\cite{langdon:2005:metas,gecco07:workshop:dcor} distributed
evolutionary computation and it has been used by several
authors ever since. More recent efforts include
\cite{duda2013distributed,DBLP:journals/corr/abs-0801-1210,EvoStar2014:jsEO,martinez2015capataz,pan2015gray}. In fact, this last paper \cite{pan2015gray} performs an analysis of what is called {\em Gray computing} covering aspects of
feasibility, cost-effectiveness, change in the user\'s experience and the
architectural optimization needed, concluding that the computing power
available is vast and it can be cost-effective to use it.

The SPACE framework \cite{leclerc2016seamless}, using a peer to peer approach, distributes fitness evaluations across a heterogeneous pool of cloud compute nodes and volunteer computers running a browser. In SPACE, peers establish a bi-directional communication with the server using the \texttt{Socket.io} JavaScript library.
As an experiment they provide an \texttt{ASM.js} compiled version of the RoboGen software platform, showing that JavaScript can be used in a broad
spectrum of applications.

Recent works have been using crowdsourcing in order to train robots
for human interaction, a representative work of this trend is that of
Anetsberger \& Bongard \cite{anetsbergerrobots} in which they propose
training simulated robots for the grounding of language symbols. They do this by using human observers that issue arbitrary commands to these robots via the web, providing positive or negative reinforcement to the resulting actions of the robot. For interaction, they use the Twitch streaming platform where users can issue commands through the
platform\'s integrated chat service. Subjects were incentivized to
interact with the system by GUI features that provided participants
with a sense of involvement with the simulation. This kind of system
could add additional dynamics, because users can collaborate or
even compete with each other when issuing their commands. This approach has been called {\em social cloud} \cite{6404452}, emphasizing the fact that user participation is gamified in order to obtain the most from it \cite{7027564}.

The proof-of-concept efforts described above, do not go any further than trying to find out how many users join the effort, and how many of them the system can support. In fact, systems such as the one described in
\cite{gecco07:workshop:dcor} had severe scaling issues; some of them
also tried to find out how much time was needed to find the solution
or, alternatively, how many users would be needed to be competitive
against single-user single-computer implementations of the same
algorithm. Lately, researchers have tried to integrate volunteer computing techinques as a part of a larger distributed computing effort \cite{leclerc2016seamless}.  In fact, systems using volunteers exclusively, exhibit a certain amount of unpredictability, and they might be better used in combination with ready-available computing power.

As indicated in the introduction, the human is an integral part of the system, which we can consider a human computation {\em human computation} \cite{quinn2011human}. Giving the user more control, and adding visibility to their participation might have positive effects. In fact, the BOINC project applies this technique by having a web page presenting the ``Top 100 multi-project BOINC participants'' where the name of the volunteer, the number of projects, GFLOPS, country and team are displayed
(\url{https://boinc.berkeley.edu/chart_list.php}).

@@

\section{Gamification}
\label{sec:gamification}
A definition given by Huotari  \cite{huotari2012defining} is ``Gamification is
the process by which gaming concepts are brought to real world tasks associated with
real people''. Gamification uses game design elements out of the domain of games
with the objective of enhancing the user\'s experience, engagement, productivity,
learning, among others. Deterding et al. proposes the following definition:
 ``Gamification” is the use of game design elements in non-game contexts''
 \cite{deterding2011gamification}.

Gamification techniques in a volunteer context seeks to persuade
users to use their natural desire to compete, learn and socialize in
given non-game context application \cite{deterding2011game,hamari2014does}.
Some works give a form of reward to users, these include
points \cite{sutter2010browse}, achievement badges or levels \cite{hamari2011framework},
the filling of a progress bar \cite{o2010get}, or providing the user with virtual currency.
By Making the rewards for  tasks achievements visible to other players or
providing leader boards are ways of encouraging players to compete \cite{hickman2010total}. In this work, a leader board was implemented in order to promote competition, presenting only the all-time top five participating users. Competition can also have problematic consequences, which can result in
negative conduct, low cooperation and collaboration, or disadvantaging certain player demographics
such as women \cite{kumar2013gamification}. Another gamification techinque
is to make existing tasks feel more like games \cite{deterding2010just}.
Some techniques used in this approach include adding meaningful choice,
on-boarding with a tutorial, increasing challenge, and adding narrative \cite{mcgonigal2011reality}.

@@

In this experiment a gamification technique is applied to
the {\sf NodIO} volunteer computing framework, using the
particular version described in \cite{DBLP:conf/gecco/MereloCGCRV16,2016arXiv160101607M}.
The objective of the experiment is to test the kind of impact
applying a gamification technique has on user engagement in this
particular application. In order to apply a rewarding system
user authentication had to be developed first. In earlier designs
this functionality was not desired because it was seen as a barrier
for participation. After all, the advantage of a browser based volunteer system
is precisely the minimum amount of user intervention needed to start.
The last thing a user wants to see is yet another registration form.
Then, the first design decision for this version is that registration is optional and simple as shown in Figure \ref{fig:login}.

@@

After login, registered users are welcomed by their name, and they can see
their score and level. All users can see in a modal dialog the leader board.
While the modal is opened, the scores are refreshed every second. All other
functionality is available to all users and is the same as the previous
{\sf NodIO} version, showing the state of the current algorithm.
In the front page a link to the open source code of the experiment was also
available \url{https://undisclos.ed}
% \url{https://github.com/lucero21/login-master}

\subsection{Problem}
In the browser, each page visiting the experiment loads an HTTP Web Worker
that runs a local island of an evolutionary algorithm to solve a
multi-modal problem called {\em l-trap} which has been used extensively
as a benchmark for evolutionary algorithms and is represented in figure \ref{fig:trap},  \cite{fernandes2009using,nijssen2003analysis}.
This function counts the number of bits in a sequence $l$ and assigns
the local maximum $a$ if it has got 0 bits and the global maximum $b$ if it has $l$
bits, this makes the fitness fall into a {\em trap}
as the number of bits is increased, decreasing linearly until a change in slope
is reached at point $z$, adding a deceptive component for evolutionary algorithms. Trap functions can be concatenated in order to increase the difficulty . In our case we have used $40$ concatenated traps. The trap function is defined as:
\[ f(u)=
    \begin{cases}
      \frac{a}{z}(z-u) & \text{if } u\leq z\\
      \frac{b}{l-z} (u-z)& \text{otherwise}
   \end{cases}
 \]
 %
\begin{figure*}[htb]
    \centering
        \includegraphics[width=2.5in]{img/trap.png}
    \caption{Trap function.
    }
    \label{fig:trap}
\end{figure*}

@@

The experiment was planed as a usibility test, employing a relativly small sample of users.

A/B testing or split testing is an experimental approach to web desig

@@

The objective of this work was to test the kind of impact
applying a gamification technique has on user engagement in
the {\sf NodIO} volunteer computing framework; In concordance with
the results obtained in other
browser-based volunteer systems, after
applying the gamification techniques, user participation was increased.
